BlueTracer: \\ a Robust API Tracer for Evasive Malware

### 2018-07-07 introduction.tex
Added Improved Introduction

Malicious software (or malware) is any software specifically designed to bring harm to a computer system. The problem posed by malware is one which is becoming increasingly important, as new and more sophisticated malware samples arise every day and the economical damage for organizations keeps worsening \cite{Cisco}. To face this threat, professionals are typically aided by a range of automatic tools capable of analysing and detecting malicious software.

Malware analysis can be carried out either statically or dynamically. Dynamic analysis encompasses techniques that execute a sample and observe the actions it actually performs, whereas in static analysis the sample is examined without running it. Such techniques have evolved over time to keep track with the increasing complexity and diversity of malware. However, in recent years, the shift towards automation, caused by the need of dealing with a huge and ever-growing number of samples, together with the rising complexity of obfuscation mechanisms utilized by malware, has strongly favoured dynamic analysis.

One of the most employed dynamic analysis techniques is function call monitoring. Generally, a function is made up of code which carries out a particular task, like for example creating a file or printing a message. Although the utilization of functions allows for easy re-usability of code and simpler maintenance, the propriety which makes them particularly valuable from a program analysis perspective is that they abstract the implementation details, providing a semantically richer representation of some functionality. For instance, let us consider a sorting function; it might not be important to know the underlying sorting algorithm as long as it is known that the function sorts the input number set. In the context of dynamic analysis, the abstractions provided by API calls and system calls (or eventually Windows Native APIs) are incredibly helpful since they can be used to grasp the overall behaviour of the sample being analyzed.

The typical technique used for function calls monitoring in dynamic malware analysis is \textit{API hooking}, i.e. the interception of function calls provided by DLLs. The idea is to alter the original sample so that, besides the function of interest, a \textit{hooking} function is also called, which is in charge of performing the wanted analysis, e.g. logging the function invocation on a file or analyzing the function's parameters \cite{Egele:2008:SAD:2089125.2089126}.

A problem that all dynamic analysis techniques have to face, including function call monitoring by means of API hooking, is the widespread of evasive malware. The malware samples belonging to this category check whether or not they are being executed in an adverse environment and conceal their harmful behaviours accordingly, like for example by carrying out an exit sequence \cite{BluePill}. Evasive techniques are frequently adopted my malicious samples. According to Symantec's Internet Security Report of March 2018, 18\% of new malware were virtual-machine-aware \cite{Symantec}. To make matters worse, the API hooking techniques presented in literature are easily detectable and are not coupled by any mechanism to hide their presence from evasive malware.

\paragraph{Contributions.}
Throughout this thesis we present \textbf{BlueTracer}, a robust library and system call tracer for Windows applications, specialized in the monitoring of evasive malware. BlueTracer is based on the Intel Pin \cite{Pin} dynamic binary instrumentation (DBI) framework and is able to counteract malware's evasion measures thanks to its integration with BluePill, a software toolkit built on top of a DBI layer which allows the simulation of the execution environment a particular malware sample was designed for and conceals any virtualization artifacts and setup details which might set off evasion \cite{BluePill}. BlueTracer is capable of tracing the input values, the output values and the return values of an extremely wide range of system calls (including Windows Native APIs) and API calls. Moreover, it also supports the tracing of Windows callbacks functions and Windows asynchronous procedure calls (APC). The tool was tested on a benign application aimed at assessing how good an anti-malware system is against evasion techniques and on actual evasive malware, proving to be effective in both tracing the samples' activity and remaining undetected.

>>>>

The term malicious software (\textit{malware} for short) describes any software specifically designed to bring harm to a computer system. Accurate and efficient analysis of malware is a compelling problem that is becoming increasingly important as hundreds of new malware strains arise every day in an ever-growing trend, while the economical damage for organizations keeps worsening~\cite{Cisco}.

To face this threat, professionals are typically aided by a range of automatic tools capable of analyzing and detecting malicious software. In particular, malware analysis can be carried out either statically or dynamically. Static analysis involves the inspection of the different code and data sections of a binary, but is often frustrated by obfuscation mechanisms and other countermeasures employed by malware authors. Dynamic analysis techniques dodge the problem by executing a malware sample and observing the actions it performs on the environment, and are better suited for dealing with large quantities of samples; these two factors have strongly favored the adoption of dynamic analyses in the security practice.

One of the most employed dynamic analysis techniques is function call monitoring. Generally speaking, a function is made up of code that carries out a particular task, such as creating a file or printing a message. A property that makes functions particularly valuable from a program analysis perspective is that they can abstract implementation details - especially in the case of library and system calls - providing a semantically richer representation of some functionality. Consider for instance a sorting function: details of the underlying algorithm might not be important as long as one knows that the function sorts an input list. In the context of dynamic analysis, the abstractions embodied by API calls and system calls are incredibly helpful since they can be used to grasp the visible behavior of the sample under analysis, regardless of any obfuscations made on its code.

The most popular technique used for function call monitoring in dynamic malware analysis is API hooking, i.e., the interception of function calls provided by dynamically linked libraries (DLL). The idea is to alter execution so that for each call in the sample, besides the function of interest, a hooking function is called, too. Such function is in charge of performing the desired analysis, which may require for instance logging the invocation on a file or analyzing its parameters.

A problem that nearly all dynamic analyses - including API hooking - have to face is adversarial behavior from \textit{evasive malware}, i.e., malware that active checks whether it is being executed in a hostile environment, such as a well-known sandbox solution for dynamic analysis. In that case, an evasive sample conceals its harmful behavior to avoid detection, for instance by executing an exit sequence. Evasive techniques are frequently adopted by modern malware. According to a March 2018 Symantec Internet Security Report, 18\% of recently observed new samples are aware of hardware virtualization solutions used also for malware analysis \cite{Symantec}. To make matters worse, most API hooking techniques proposed to date in literature are easily detectable by malware, and not coupled with mechanisms to hide their presence from evasive malware.

\paragraph{Contributions.}The core contribution of my thesis is the design and implementation of BlueTracer, a robust library and system call tracer for Windows programs specialized in evasive malware.

BlueTracer is based on the Intel Pin~\cite{Luk05} dynamic binary instrumentation framework, which allows for transparently injecting analysis code in a binary program as it executes. BlueTracer can effectively counteract a large variety of evasive techniques due to its integration with BluePill~\cite{BluePill}, a stealthy execution framework being developed at Sapienza University and Royal Holloway University of London that allows the simulation of the original execution environment a particular malware sample was designed for, and conceals any analysis artifacts or setup details that might set off evasion \cite{BluePill}.

BlueTracer can trace input parameters, output buffers and return values of an extremely wide range of system calls (including Windows Native APIs that are invoked directly via special CPU instructions in more advanced malware) and library calls. Moreover, it also supports the tracing of user-mode callbacks and Windows Asynchronous Procedure Calls.

The heterogeneity of Windows libraries used in malware and the lack of well-structured documentation for their prototypes and input/output parameters posed a significant implementation challenge, which BlueTracer addresses by leveling out and integrating reliable external sources (two industry projects: Dr. Memory and CISCO PyREBox) that account for more than 17,000 existing function prototypes. An important engineering effort was also required to make best use of Intel Pin’s capabilities with respect to run-time CPU and memory costs, as well as to address some inherent limitations that emerged on the field.

We first validated BlueTracer on benign applications exercising a large number of functions, and then on a popular security product used for assessing how stealthy a malware analysis system is against a large body of public evasion techniques used in the wild. The obtained results suggested that BlueTracer can be very effective in tracing a malware sample's activities while remaining undetected. 
\iffalse
I then analyzed a set of highly evasive samples collected by Joe Security, the proclaimed technology leader for the analysis of evasive malware, as they feature exotic evasions that most analysis products have struggled to deal with. The logs collected by BlueTracer reveal behaviors consistent with the analysis reports authored by Joe Security, suggesting that BlueTracer can be very effective in tracing a malware sample's activities while remaining undetected. 
\fi
